---
title: "FACTscorer R Package"
output: 
  github_document:
    toc: true
    toc_depth: 4
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## FACTscorer Summary

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/FACTscorer)](https://CRAN.R-project.org/package=FACTscorer)
<!-- badges: end -->

`FACTscorer` contains functions to score most of the patient-reported outcome
(PRO) measures from the Functional Assessment of Cancer Therapy (FACT) and
Functional Assessment of Chronic Illness Therapy (FACIT) family of PRO measures.
The questionnaires themselves can be downloaded from the [FACIT website](https://www.facit.org/).  For most of
the FACT/FACIT questionnaires, the [FACIT website](https://www.facit.org/) provides scoring syntax for use with
commercial statistical software (SAS and SPSS).  The `FACTscorer` R package is
intended to serve as a free, reliable alternative for those without access to
SAS or SPSS.  Additionally, it enables R users to both score and analyze the
FACT and FACIT scales in R, avoiding the time-consuming and error-prone process
of transferring data back-and-forth between statistical software. Finally, use
of the `FACTscorer` package will prevent many sources of scoring error common when
using SAS and/or SPSS syntax (e.g., copy-paste errors and other accidental
modifications to the syntax).

## Installation

You can install the development version of `FACTscorer` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("raybaser/FACTscorer")
```

## Expected Input Data Format

+ **All `FACTscorer` functions expect the FACT/FACIT items to be named according
to FACT/FACIT conventions.**  For example, for the FACT-G, the first physical
well-being item should be named GP1, the second GP2, and so on.  Please refer to
the materials provided on the [FACIT website](https://www.facit.org/) for the
particular questionnaire you are using.  In particular, refer to the left margin
of the official questionnaire (i.e., from the [FACIT website](https://www.facit.org/)) 
for the appropriate item variable names.  

+ All items on the questionnaire you are trying to score must be included as variables (**appropriately named**, see above) in your data frame.  

+ All item variables must be numeric or integer class (not character or factor) and be whole numbers ranging from 0 to 4.  

  + For consistency with the SAS and SPSS scoring syntax on the [FACIT website](https://www.facit.org/), item values of `8` and `9` are allowed but will be recoded as system missing (i.e., `NA`) before scoring.



## Examples  

### Scoring the FACT-G

This example shows how to score the FACT-G (Functional Assessment of Cancer Therapy - General) using the `scoreFACTG` function.

Given a data frame that includes all 27 of the FACT-G items as variables,
**appropriately named and formatted**, this function generates all of the FACT-G scale
scores.  


```{r FACT_G}
library(FACTscorer)

# Use `make_FACTdata()` function to create a small data frame with fake FACT-G
# item data to score.  
##  - Each respondent will have roughly 10% of the items missing to illustrate the
##    scoring function properly prorates the scores for missing items.  
##  - See ?make_FACTdata for more info on the arguments, if desired.

df_factg <- make_FACTdata(N = 8, pmiss1 = .10, pmiss2 = .10)

# Score the FACT-G items using the `scoreFACTG` function
# All FACT-G scale scores will be output to a new data frame named `df_factg_scores`
df_factg_scores <- scoreFACTG(df_factg)
df_factg_scores

# There are several ways you can merge the scores back into your original data
# frame.  For example:
df_factg <- cbind(df_factg, df_factg_scores)
df_factg

```


### Scoring the FACT-B (Breast Cancer)


This example shows how to score one of the cancer-specific measures in the FACT family, the FACT-B (Functional Assessment of Cancer Therapy - Breast).  The FACT-B can be scored using the `scoreFACT_B` function.

Given a data frame that includes all of the FACT-B items as variables,
**appropriately named and formatted**, the `scoreFACT_B` function generates all of the FACT-B scale scores.  

**NOTE**:  The FACT-B contains all of the FACT-G (General) items, plus 10 items that assess "Additional Concerns" specific to breast cancer patients.  Those 10 items are used to calculate the Breast Cancer Subscale (BCS).  



```{r FACT_B}
library(FACTscorer)

# Use `make_FACTdata()` function to create a small data frame with fake FACT-B
# item data to score.  
## - The `namesAC` argument takes a character vector containing the names of  
##   the 10 "Additional Concerns" (AC) items. 
df_factb <- make_FACTdata(N = 8, pmiss1 = 0.10, pmiss2 = 0.10,
  namesAC = c('B1', 'B2', 'B3', 'B4', 'B5', 'B6', 'B7', 'B8', 'B9', 'P2'))

# Score the FACT-B items using the `scoreFACT_B` function
# All FACT-B scale scores will be output to a new data frame named `df_factb_scores`
df_factb_scores <- scoreFACT_B(df_factb)
df_factb_scores

# There are several ways you can merge the scores back into your original data
# frame.  For example:
df_factb <- cbind(df_factb, df_factb_scores)

```



### Scoring only "Additional Concerns" items from the FACT-B

+  The **full** FACT-B contains all 27 of the FACT-G (General) items, plus 10
items that assess "Additional Concerns" specific to breast cancer patients. 
Those 10 "Additional Concerns" (AC) items are used to calculate the Breast
Cancer Subscale (BCS).  

+  Although not generally recommended by the FACIT organization, sometimes researchers might omit the FACT-G items and include only the breast cancer-specific "Additional Concerns" items in their study.   

+  Functions in the `FACTscorer` package can be told to score only the "Additional Concerns" items using the argument, `AConly = TRUE`.  In the case of the FACT-B, using the  `scoreFACT-B` fucntion with `AConly = TRUE` will score only the Breast Cancer Subscale (BCS).   

+  Although the example below uses the FACT-B as an example, this applies to all of the FACT/FACIT questionnaires that include the general FACT-G items plus more specific "Additional Concerns" items.   



```{r FACT_B_AConly}
library(FACTscorer)

# Use `make_FACTdata()` function to create a small data frame with fake FACT-B
# item data to score.  
##  - The `namesAC` argument takes a character vector containing the names of  
##    the 10 "Additional Concerns" (AC) items.  
##  - Setting `AConly = TRUE` will generate data for the AC items and omit the FACT-G
##    items.  
df_bcs <- make_FACTdata(N = 8, pmiss1 = 0.10, pmiss2 = 0.10, AConly = TRUE,
  namesAC = c('B1', 'B2', 'B3', 'B4', 'B5', 'B6', 'B7', 'B8', 'B9', 'P2'))

# Use the `scoreFACT_B` function with `AConly = TRUE` to score only the "Additional Concerns" items.  
# The data frame `df_bcs_scores` will contain only the Breast Cancer Subscale (BCS) scores.
df_bcs_scores <- scoreFACT_B(df_bcs, AConly = TRUE)
df_bcs_scores

# There are several ways you can merge the scores back into your original data
# frame.  For example:
df_bcs <- cbind(df_bcs, df_bcs_scores)


```



